Machine Learning guided workflow for Ribosome Binding Site engineering

### 2021-04-26 paper/main.tex
Updated paper - redid according to Nature Comms requirments, updated Supplementary.

\section{Abstract}

Fine control of gene expression is extremely important for engineered biological systems.
Such control can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
Unfortunately, our understanding of how RBSs perform in vivo is insufficiently well understood to allow predictable, reliable design of RBSs at the level of finesse needed for some applications.
To address this problem, we have created a machine learning (ML) enabled workflow for the design of bacterial RBSs.
We used Gaussian Process Regression for prediction of relative transcription initiation rates and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
The Bandit algorithm addresses the exploration-exploitation trade-off, balancing the search for highly novel sequences with the desire for strong RBSs.
We have integrated the ML algorithms with laboratory automation and high-throughput processes, creating a robust workflow for the design of custom RBSs.
By employing these techniques, we were able to increase the reliability and reproducibility of results and increase the confidence in the design process.
Using our workflow, we generated a novel library of diverse RBSs with a wide range of expression levels.
Notably, a high number of these sites demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs.
Additionally, this work elucidated some design guidelines, including the favourable level of difference between two sequences and efficient numerical representation of the analysed DNA/RNA sequence.
In summary, by employing both machine learning and high-throughput laboratory methods we have created a workflow for creation of small genetic devices enabling efficient generation of parts with required characteristics.
As a next step, we hope to expand this workflow to more complicated parts including promoters and terminators.

@@

\appendix
\input{appendix}
\end{document}

>>>>

\section*{Abstract}

Fine control of gene expression can be achieved through engineering transcriptional and translation control elements, including the Ribosome Binding Site (RBS).
Unfortunately, RBSs are not understood at the level of finesse required for reliable design. 
To address this problem, we have created a machine learning (ML) enabled workflow for the design of bacterial RBSs.
We used Gaussian Process Regression for prediction and the Upper Confidence Bound-based Bandit algorithm for recommendation of genetic designs to be tested in vitro.
We have integrated the ML algorithms with laboratory automation and high-throughput processes, creating a robust workflow for the design of custom RBSs.
Using our workflow, we generated a novel library of diverse RBSs with a wide range of expression levels.
Notably, a high number of these sites demonstrate translation initiation rates equalling or exceeding the currently known strong RBSs.
Additionally, this work elucidated some important RBS design guidelines.

@@

\appendix
\input{supplementary}
\end{document}

### 2021-04-26 paper/supplementary.tex
Updated paper - redid according to Nature Comms requirments, updated Supplementary.

\section*{Appendix}

\section{Reproducible Plots}
In this section, we put reproducible plots and will update the plots when we get full results.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.4]{plots/swarmplots.pdf}
    \caption{Swarmplots for different groups, with raw (left) and normalised (right) TIR labels (averaged over 6 replicates). 
    Group names represent: Consensus (consensus sequence tested in different round); BPS-NC (bps noncore); BPS-C (bps core); UNI (uniformly random); PPM (position-based probability matrix); Bandit-0 (bandit design for round 0); Bandit-1 (bandit design for round 1).}
    \label{fig:swarmplots.}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.3]{plots/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
    \includegraphics[scale = 0.3]{plots/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
    \caption{Kernel Heatmap and Predictions. Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265). Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right). The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
    \label{fig:my_label}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.25]{plots/clustering.png}
    \caption{TSNE of RBS sequences in design space with clustering. The distance is calculated based on the weighted degree kernel on RBS sequences. Colours indicate different groups, shapes indicate different clusters.}
    \label{fig:my_label}
\end{figure}

\section{Gaussian Process Regression}

@@

\section{Choices of Kernels}

@@

\subsection{Normalisation of Kernel}

@@

\section{Batch Recommendation}

@@

\section{Design Pipeline}

@@

\section{Core vs non-core regions of the RBS}
According to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS. 
This has also been seen in this work in the 1by1 group of RBSes. Change in base in the core region has had a much more profound effect on the TIR than when the base change happened outside the core (Figure \ref{fig:core_vs_noncore}). The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.6]{plots/core_vs_noncore.pdf}
    \caption{\textbf{Comaprison of base change impact on TIR in core versus non-core region.} The core region is highligthed in light green and the lines are rolling averages for each base. The TIRs obtained for RBSes where the base change occured outside the core region are significantly less }
    \label{fig:core_vs_noncore}
\end{figure}

\section{Intuition behind UCB and visualisation}

@@

\section{Result analysis}

@@

\section{Figures}

>>>>

\textbf{Supplementary}

\section{Extended Machine Learning Discussion}

\subsection{Reproducible Plots}
In this section, we put reproducible plots and will update the plots when we get full results.

\begin{figure}[h]
    \centering
    \includegraphics[scale = 0.4]{plots/Supplementary/swarmplots.pdf}
    \caption{Swarmplots for different groups, with raw (left) and normalised (right) TIR labels (averaged over 6 replicates). 
    Group names represent: Consensus (consensus sequence tested in different round); BPS-NC (bps noncore); BPS-C (bps core); UNI (uniformly random); PPM (position-based probability matrix); Bandit-0 (bandit design for round 0); Bandit-1 (bandit design for round 1).}
    \label{fig:swarmplots.}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_TIR_Label.pdf}
    \includegraphics[scale = 0.3]{plots/Supplementary/Data_Similarity_and_Prediction_Sorted_by_RBS_Similarity.pdf}
    \caption{Kernel Heatmap and Predictions. Sequences are grouped as Consensus (1-2), BPS-C (3-20), BPS-NC (21-61), UNI (62-90), PPM (91-118), Bandit-0 (119-177), Bandit-1 (178-265). Inside of each group, sequences are clustered and sorted in terms of TIR labels (left) or RBS similarity (right). The first row shows the similarity measured by weighted degree kernel with shift, the second shows the predicted mean and uncertainty (1.95 standard deviation).}
    \label{fig:my_label}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale = 0.25]{plots/Supplementary/clustering.png}
    \caption{TSNE of RBS sequences in design space with clustering. The distance is calculated based on the weighted degree kernel on RBS sequences. Colours indicate different groups, shapes indicate different clusters.}
    \label{fig:my_label}
\end{figure}

\subsection{Gaussian Process Regression}

@@

\subsection{Choices of Kernels}

@@

\subsubsection{Normalisation of Kernel}

@@

\subsection{Batch Recommendation}

@@

\subsection{Design Pipeline}

@@

\subsection{Intuition behind UCB and visualisation}

@@

\subsection{Result analysis}

@@

\subsection{Statistics of the core vs non-core regions of the RBS}

In our study, we have tested a set of sequnces designed to confirm the notion that changing bases within the RBS core (6 bases) is statistically more influencing the TIR than the changes made outside the core.
This hypothesis has been build based on reported biases towards certain bases present in the core of the RBS but absent outside of it.
For example, according to \cite{SHULTZABERGER2001} there is a strong bias towards A and G bases in the core region of the RBS.
Similarly outside of the 6 bases of the core in the wider 20 bp context of the RBS there is no significant bias towards any particular base which suggest that these bases do not contribute to the overal TIR of a given RBS. 
This effect is shown in (Figure \ref{fig:core_vs_noncore}), which shows results for our set X of sequences.
The value of Welch's t-test between the mean TIR in core and non-core groups is -4.8780 with p-value < 0.0001 and 34 degrees of Freedom.

\begin{figure}[h]
    \centering
    \includegraphics[scale=0.4]{plots/Supplementary/core_vs_noncore.pdf}
    \caption{\textbf{Comparison of base change impact on TIR in core versus non-core region.} The core region is highligthed in light green and the lines are rolling averages for each base. The top dotted line shows the TIR for the benchmark sequence, where dots represent a change at a given position to a given base, which is colour coded.}
    \label{fig:core_vs_noncore}
\end{figure}

\section{Supplementary Figures}

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \includegraphics[scale=0.25]{plots/Supplementary/Structure GGGGGG.pdf}
         \caption{$y=x$}
         \label{fig:GGGGGG}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \includegraphics[scale=0.25]{plots/Supplementary/Structure GGGGGC.pdf}
         \caption{$y=3sinx$}
         \label{fig:GGGGGC}
     \end{subfigure}
     \vskip\baselineskip
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \includegraphics[scale=0.25]{plots/Supplementary/Structure CGGGAC.pdf}
         \caption{$y=5/x$}
         \label{fig:CGGGAC}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.49\textwidth}
         \centering
         \includegraphics[scale=0.25]{plots/Supplementary/Structure CACGTT.pdf}
         \caption{$y=5/x$}
         \label{fig:CACGTT}
     \end{subfigure}
        \caption{\textbf{Folding predictionds for four different RBS.} Here we show the predicted, energetically most favourable structures for four of our RBS with the following cores: GGGGGG, GGGGGC, CGGGAC, CACGTT and their immediate upstream and downstream background sequence. There is no discerbile and consistent difference between strong ones (the first two) and weak ones (the last two).}
        \label{fig:structures}
\end{figure}
